---
title: "Projected Energy Burden — 2024 Estimates"
subtitle: "Georgia Power Service Territory"
author: "Energy Equity Project"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  word_document:
    toc: false
    fig_caption: true
    reference_docx: null
knit: (function(inputFile, encoding) {
    rmarkdown::render(
      inputFile,
      encoding = encoding,
      output_dir = "../outputs",
      output_file = paste0(format(Sys.Date(), "%d-%m-%Y"), "-projected_burden_brief.docx")
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 6.5,
  fig.height = 4,
  dpi        = 300,
  out.width  = "100%"
)

library(tidyverse)
library(scales)
library(knitr)
library(flextable)
library(glue)

# Helper: find most recent CSV matching glob pattern
find_output <- function(pattern) {
  f <- tail(sort(Sys.glob(file.path("../outputs", pattern))), 1)
  if (length(f) > 0) read.csv(f) else stop(glue("No file found matching: {pattern}"))
}

# Read CSVs
burden_by_fpl  <- find_output("*-lead_burden_by_fpl_projected_2024.csv")
heag_total     <- find_output("*-lead_heag_total_projected_2024.csv")
burden_by_race <- find_output("*-lead_burden_by_race.csv")

# Territory-wide overall weighted average burden (back-calculated from FPL tiers)
overall_burden <- sum(burden_by_fpl$wgt_mean_burden * burden_by_fpl$total_units) /
                  sum(burden_by_fpl$total_units)

total_hh <- sum(burden_by_fpl$total_units)

# HEAG scalars
hh_above_6   <- heag_total$households_above_6pct
total_gap_m  <- heag_total$total_gap_annual_usd / 1e6
avg_gap_hh   <- heag_total$avg_gap_per_hh_annual

# Race scalars
bipoc_burden  <- burden_by_race %>% filter(racial_majority == "Majority BIPOC") %>% pull(wgt_mean_burden)
white_burden  <- burden_by_race %>% filter(racial_majority == "Majority white")  %>% pull(wgt_mean_burden)
bipoc_above_6 <- burden_by_race %>% filter(racial_majority == "Majority BIPOC") %>% pull(pct_above_6)
white_above_6 <- burden_by_race %>% filter(racial_majority == "Majority white")  %>% pull(pct_above_6)

# flextable defaults for Word compatibility
set_flextable_defaults(
  font.size = 10,
  theme_fun = theme_booktabs,
  big.mark  = ","
)
```

## 1. Territory-Wide Overview

Across Georgia Power's residential service territory, the projected 2024 weighted mean energy burden is **`r round(overall_burden, 1)`%** of household income. This estimate reflects approximately `r formatC(round(total_hh / 1e6, 2), format = "f", digits = 2)` million households in the IOU service territory. Approximately `r formatC(round(hh_above_6), format = "d", big.mark = ",")` households — those spending more than 6% of income on energy — face a combined annual Home Energy Affordability Gap (HEAG) of **`r dollar(heag_total$total_gap_annual_usd, accuracy = 1e6, scale = 1/1e6, suffix = "M")`**, with an average per-household gap of **`r dollar(avg_gap_hh, accuracy = 1)`**.

These figures are projected from the DOE LEAD 2022 baseline using EIA electricity rate growth (2022–2024) and ACS median household income growth. The 2022 baseline is also available for comparison; the projection accounts for Georgia Power's rate increases over this period.

---

## 2. Energy Burden by Income Level

Energy burden is highly concentrated among low-income households. As shown in Table 1, households at 0–100% of the federal poverty level (FPL) carry a weighted mean burden of `r round(burden_by_fpl %>% filter(fpl150 == "0-100%") %>% pull(wgt_mean_burden), 1)`%, with `r round(burden_by_fpl %>% filter(fpl150 == "0-100%") %>% pull(pct_above_6), 0)`% of households in this tier exceeding the 6% affordability threshold. Burden falls sharply at higher income tiers — households at 200–400% FPL carry a mean burden of `r round(burden_by_fpl %>% filter(fpl150 == "200-400%") %>% pull(wgt_mean_burden), 1)`%, with only `r round(burden_by_fpl %>% filter(fpl150 == "200-400%") %>% pull(pct_above_6), 1)`% above the threshold. This gradient reflects that energy costs are largely fixed while incomes vary — making energy a larger share of the budget for lower-income households regardless of usage.

**Table 1. Projected 2024 Energy Burden by Income Level (FPL Tier)**

```{r table-burden-by-fpl}
burden_by_fpl %>%
  mutate(
    wgt_mean_burden = round(wgt_mean_burden, 1),
    median_burden   = round(median_burden, 1),
    total_units     = round(total_units, 0),
    pct_above_6     = round(pct_above_6, 1)
  ) %>%
  rename(
    `FPL tier`             = fpl150,
    `Wgt. mean burden (%)` = wgt_mean_burden,
    `Median burden (%)`    = median_burden,
    `Households`           = total_units,
    `% above 6% threshold` = pct_above_6
  ) %>%
  flextable() %>%
  colformat_num(j = "Households", big.mark = ",", digits = 0) %>%
  set_table_properties(width = 1, layout = "autofit") %>%
  bold(part = "header")
```

*Source: DOE LEAD v4 (2022), projected to 2024. Burden = estimated annual energy cost ÷ estimated annual household income. FPL tiers derived from LEAD income band crosswalk.*

---

## 3. Energy Burden by Racial Composition

Majority-BIPOC census tracts carry higher aggregate energy burdens than majority-white tracts, reflecting differences in income composition rather than differential energy pricing:

- **Majority-BIPOC tracts**: weighted mean burden of `r round(bipoc_burden, 1)`%; `r round(bipoc_above_6, 0)`% of households above the 6% threshold
- **Majority-white tracts**: weighted mean burden of `r round(white_burden, 1)`%; `r round(white_above_6, 0)`% of households above the 6% threshold

The `r round(bipoc_burden - white_burden, 1)` percentage-point gap between tract types is driven by income distribution: majority-BIPOC tracts are more heavily concentrated at lower FPL tiers, where burdens are highest for all households. Within the same income bracket, burdens are nearly identical across tract types — the disparity is structural, not a function of race per se. A companion brief (03b) provides full detail on the racial disparity analysis.

---

## 4. Data and Methodology Note

**Projection method.** DOE LEAD v4 (2022) provides the baseline energy burden estimates at the census-tract level. Projected 2024 figures adjust energy cost estimates by the ratio of Georgia's average retail electricity rate in 2024 to 2022 (source: EIA Form 861), and adjust income estimates by the ratio of Georgia's ACS median household income in 2024 to 2022. Both multipliers are applied uniformly across all tracts.

**Baseline also available.** The 2022 DOE LEAD baseline (unadjusted) is available in `outputs/` for comparison. The projection increases burden estimates modestly, reflecting Georgia Power's rate increases over this period.

**Service territory scope.** All figures are restricted to census tracts within Georgia Power's IOU service territory, identified via the HFLID/ORNL electric retail service territory shapefile (2023 vintage; EIA utility ID 7140). Georgia has numerous electric cooperatives and municipal utilities; this analysis covers only Georgia Power's residential customers.

**HEAG definition.** The Home Energy Affordability Gap (HEAG) is the aggregate dollar amount by which households spending more than 6% of income on energy exceed that 6% threshold. It measures the scale of unmet affordability need, not total energy spending.
